Optimizing Structural Properties Of Neural Networks With Genetic Algorithms

### 2015-08-07 paper/paper.tex
paper

Our concrete goal for this article was to examine which choices of genotype
parameters are optimal for learning sinusoidal waves with relatively high
or low frequencies. 
Based on preliminary simulation results we chose
$\nu\in[0.3, 3]$ as frequency range for the challenge $C\ix{high}$, producing
high frequency waves, and $\nu\in[0.004, 0.04]$ for $C\ix{low}$, producing low
frequency waves. 
Optimizing the fitness functions $F_{C\ix{high}}$ or $F_{C\ix{low}}$ thus
results in genotypes with the desired frequency properties.

@@

A measure that is time-shift independent and maximized for functions of
identical shape is the maximum of the cross correlation 
\begin{align}
    f\ix{cor}(z, z') = \max_t \frac{\left<z, z'(\cdot
    - t)\right>}{\sqrt{\left< z, z\right> \left<z', z' \right>}}~.
\end{align}
Algorithmically, the signals $z$ and $z'$ are split into overlapping chunks
which are correlated separately.  This saves computational power and
prevents minute differences in frequency to accumulate into a huge phase
shift, so the phase only has to remain relatively constant over the
timescale of a single chunk (which is about 10 seconds of simulated time).
%This measure becomes problematic for functions that are very close to zero
%for a longer time (order of chunk size), but since we are going to work
%with superpositions of sinusoidals, this drawback
%is unproblematic.

@@

Our investigation showed that the frequency of the wave that is to be reconstructed 
strongly determines whether FORCE learning is able to learn that curve. Still, 
a good choice of network parameters is able to improve the success rate for
certain frequency ranges. Some parameters ($\ERprob$, $\feedback$) can vary in a relatively large interval
without influencing the success very much, whereas others are confined to a relatively small interval ($\gain$).

The code developed for this paper provides a framework which can be used to perform a variety of similar 
investigations, e.g. the inclusion of feed forward topologies, or the extension to task other than simple 
sinusoidals. To that end, it would be wise to improve that selection algorithm by using the same tasks
for all generators in each generation, instead of evaluating them all independently. This would make the 
selection process more robust because it removes the dependence on the task distribution.

One problem with the approach presented here is that, in order to use the results of the genetic algorithm, 
we determined a single set of parameters by taking the mean of the values of the most successful generators.
Thereby, any information about an interdependence between the parameters is lost. We checked that there are no strong
correlations between any pair of parameters by doing 3D plots that showed the two parameters for each generator 
in development with the generations and could not detect any obvious correlation, but have not investigated 
this problem thoroughly.

The biggest problem with the current implementation, however, is the large amount of "magic" numbers, 
i.e. hyper-parameters of the optimization, that are used in the code. Their values were determined
by an informal search of testing a few optimizations. However, this should mostly affect the performance
of the optimization algorithm and not the results themselves.

>>>>

Our concrete goal for this article was to examine which choice of genotype
parameters are optimal for learning sinusoidal waves with relatively high
or low frequencies. 
Based on preliminary simulation results we chose
$\nu\in[]$ as frequency range for the challenge $C\ix{high}$, producing
high frequency waves, and $\nu\in[]$ for $C\ix{low}$, producing low
frequency waves. 
Optimizing the fitness functions $F_{C\ix{high}}$ or $F_{C\ix{low}}$ thus
results in genotypes with the desired frequency properties.

@@

A measure that is time-shift independent and maximized for functions of
identical shape, on the contrary, is the maximum of the cross correlation 
\begin{align}
    f\ix{cor}(z, z') = \max_t \frac{\left<z, z'(\cdot
    - t)\right>}{\sqrt{\left< z, z\right> \left<z', z' \right>}}~.
\end{align}
Algorithmically, the signals $z$ and $z'$ are split into overlapping chunks
which are correlated separately.  This saves computational power and
prevents minute differences in frequency to accumulate into a huge phase
shift, so the phase only has to remain relatively constant over the
timescale of a single chunk (which is about 10 seconds of simulated time).
%This measure becomes problematic for functions that are very close to zero
%for a longer time (order of chunk size), but since we are going to work
%with superpositions of sinusoidals, this drawback
%is unproblematic.

@@

%Our investigation showed that the frequency of sinusoidal tasks strongly
%determines whether FORCE learning is able to learn it or not. Still, a good
%choice of network parameters, conveying structural information, is able to
%improve the success rate for certain frequency ranges. 
%Under certain conditions, some parameters (like $\ERprob$ for low
%frequencies or $\feedback$ for high frequencies) can vary in a relatively
%large interval without influencing the success critically.

While we only presented results concerning simple sinusoidals of varying
frequencies here, the code developed during our project provides
a framework which can be used to perform a variety of similar and extended
investigations; for example the inclusion of feed forward topologies and
arbitrary (higher dimensional) tasks with optional external input pathways
(e.g.\ as a timer signal to make the network synchronize better).

One peculiarity to be recognized throughout our project work is that the
optimization algorithm does not excel at providing strictly converging,
single most optimal parameter values, but that we are rather given more or
less broad ``survival ranges'' for each parameter as result. And though the
final genotypes varied quite a bit in their fitness, we did not observe
strong correlations between the different parameters in these ranges when
doing sporadic tests.

Indeed, looking at all of our simulations in their entirety, the genotype
parameters showed less influence on the learning results than we initially
expected, and during our inquiry it often seamed that the statistics of the
challenges might overshadow the effect of the parameters -- at least when
looking at more complex tasks than monofrequent waves. A wise improvement
would therefore be to make the challenges more deterministic, e.g.\ to use
the same tasks for all genotypes in each generation instead of drawing them
independently from some (wide range) distribution.

Another issue to be mentioned with the current implementation is the large
amount of empirical (``magic") numbers, i.e.\ hyper-parameters of the
optimization used in the code. Examples hereof are the strength of the
mutation process or the mechanics of the selection process (actual number
of survivors, survival chances for weaker genotypes, etc.).  Their values
were determined by an informal search of testing a few optimizations but
were not investigated thoroughly. This should, however, mostly affect the
performance of the optimization algorithm and not the results themselves.

%One problem with the approach presented here is that, in order to use the
%results of the genetic algorithm, we determined a single set of parameters
%by taking the mean of the values of the most successful generators.
%Thereby, any information about an interdependence between the parameters
%is lost. We checked that there are no strong correlations between any pair
%of parameters by doing 3D plots that showed the two parameters for each
%generator in development with the generations and could not detect any
%obvious correlation, but have not investigated this problem thoroughly.

Nevertheless *ABSCHLUSSSATZ*

%This would certainly make the selection process more robust as it removes
%the dependence of the task distributions.
%In general, the genotypes parameters showed less influence than we expected.
%, whereas others are
%restricted to more concise regions (???).
%Some parameters
%($\ERprob$, $\feedback$) can vary in a relatively large interval without
%influencing the success very much, whereas others are confined to
%a relatively small interval ($\gain$).

%To that end, it would be wise to improve that selection
%algorithm by using the same tasks
%for all generators in each generation, instead of evaluating them all independently. This would make the 
%selection process more robust because it removes the dependence on the task distribution.

%One problem with the approach presented here is that, in order to use the results of the genetic algorithm, 
%we determined a single set of parameters by taking the mean of the values of the most successful generators.
%Thereby, any information about an interdependence between the parameters is lost. We checked that there are no strong
%correlations between any pair of parameters by doing 3D plots that showed the two parameters for each generator 
%in development with the generations and could not detect any obvious correlation, but have not investigated 
%this problem thoroughly.
